Exponential Random Graph Models with Big Networks:\\ Maximum Pseudolikelihood Estimation and the Parametric Bootstrap

### 2017-07-17 TeX/conference_paper.tex
Big Data Conference paper
I have written the paper using the latex template that is required for the Big Data Conference

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={140mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage{listings}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

\begin{document}

\setcounter{section}{-1}

\renewcommand{\refname}{Bibliography}

\onehalfspacing
\setlength{\headsep}{15mm}

\thispagestyle{plain}

\title{\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap}
\maketitle

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000. When the network is large, one option for estimation is maximum pseudolikelihood (MPLE). MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks. The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel. We compare the two different approaches by applying them on U.S. Supreme Court Citation Network Data.
\end{abstract}

 \section*{Introduction}

The field of network science faces a double-edge sword when it comes to computational limitations to innovation. First, the networks under study are growing larger with each year. {\bf [CHRISTIAN, WOULD YOU CHECK OUT N NODES IN {\em SOCIAL NETWORKS}?]}. Second, analytical methods are growing more sophisticated, increasingly involving estimation and optimization, going beyond descriptive calculations {\bf [Could also classify articles based on whether a statistical model is used {\em SOCIAL NETWORKS}?]}. In order to avoid a limiting horizon in which state-of-the-art methods cannot be used with state-of-the-art ``big'' data, network methodologists need to consider all available options in easing the computational burden of tools for network analysis.

 \section*{The Exponential Random Graph Model}
The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al. \cite{robins.pattison.2007}). In the following, we will focus on directed networks; however, undirected networks can be introduced very similar.\\[0.3cm] 
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

\section*{Estimation}

@@

\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] = \frac{c(\theta)}{c(\theta_0)}
\label{konver}

@@

\text{loglik}(\theta)-\text{loglik}(\theta_0)&=&- \log(c(\theta))+\log(c(\theta_0))\\

@@

\section*{Efficiency of MPLE/MCMLE}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.5]{RMSE_10k}
\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
\label{rmse}
\end{center}
\end{figure}

@@

\section*{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough. The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.

>>>>

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
   \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

\usepackage[table,xcdraw]{xcolor}
\usepackage{multirow}
\usepackage{multicol}

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Exponential Random Graph Models with Big Networks:\\ Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations

\author{\IEEEauthorblockN{Christian S. Schmid}
\IEEEauthorblockA{Department of Statistics\\
The Pennsylvania State University\\
State College, USA\\
Email: schmid@psu.edu}
\and
\IEEEauthorblockN{Bruce A. Desmarais}
\IEEEauthorblockA{Department of Political Science\\
The Pennsylvania State University\\
State College, USA\\
Email:  bdesmarais@psu.edu}
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000. When the network is large, one option for estimation is maximum pseudolikelihood (MPLE). MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks. The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel. We compare the two different approaches by applying them on U.S. Cosponsorship Network Data.

\end{abstract}

\begin{IEEEkeywords}
network, ERGM, parametric bootstrap, maximum pseudolikelihood

\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
The field of network science faces a double-edge sword when it comes to computational limitations to innovation. First, the networks under study are growing larger with each year. {\bf [CHRISTIAN, WOULD YOU CHECK OUT N NODES IN {\em SOCIAL NETWORKS}?]}. Second, analytical methods are growing more sophisticated, increasingly involving estimation and optimization, going beyond descriptive calculations {\bf [Could also classify articles based on whether a statistical model is used {\em SOCIAL NETWORKS}?]}. In order to avoid a limiting horizon in which state-of-the-art methods cannot be used with state-of-the-art ``big'' data, network methodologists need to consider all available options in easing the computational burden of tools for network analysis.

\section{The Exponential Random Graph Model}
\noindent The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al. \cite{robins.pattison.2007}). In the following, we will focus on directed networks; however, undirected networks can be introduced very similar.\\[0.3cm] 
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

\section{Estimation}

@@

\begin{equation}\label{konver}
\begin{split}
\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right) &
~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~\\ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]& = \frac{c(\theta)}{c(\theta_0)}
\end{split}

@@

&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\

@@

\section{Efficiency of MPLE and MCMLE}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{RMSE_10k}
\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
\label{rmse}

@@

\section{Bootstrapped MPLE}
\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough. The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE. \\[0.3cm]
\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. \\[0.3cm]
For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals. Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval. 
\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
\label{coverage}
\end{figure}
